---
title: "Joins in class using FEC campaign finance data"
output: 
---

Source: [fec.gov bulk data](https://www.fec.gov/data/browse-data/?tab=bulk-data) 
Data from the 2021-2022 election cycle

#Introduction: 
Each candidate for federal office (U.S. House of Representatives, U.S. Senate, President) must register with the Federal Election Commission (FEC) and form a Political Action Committee (PAC) to receive contributions. Candidates can have several different kinds of PACs, but everyone has a "principal campaign committee" through which they receive money.

In an election (even) year: every Representative is up for re-election in an even year, since Reps serve 2-year terms. One third of the Senate is up for re-election every two years because Sens serve 6-year terms. Candidates can receive money even if they're not running for re-election, but individuals are limited in how much they can give a candidate each 2-year cycle: $2900.

Missouri has 8 House seats and 2 Senate seats; only 1 Senate seat was up for re-election in 2022; Roy Blunt decided not to run for re-election so it was an open seat. Josh Hawley wasn't up for re-election until the following election in 2024.


```{r, message = F}
library(tidyverse)

transactions <- read_csv("data/transactions.csv")
committees <- read_csv("data/committees.csv")
```

`transactions` has a lot of information about individual expenses (in this table, independent expenditures and committee donations) that are reported by particular committees. Those committees are identified in the cmte_id column; that’s the only information we have about them. To find out a committee’s name, location, affilitions, etc, you need to look in the committees file.

`committees` has one row for each PAC (political action committee) registered with the FEC; it includes the same `cmte_id` along with committee name (`cmte_nm`), location, and affiliated organizations or candidates.

Note that the tables share one column of information: `cmte_id`. They happen to be named the same thing which isn’t necessary; but they do have to contain the exact same information (in the exact same format) for the join to work.

```{r}
combined <- left_join(transactions, committees, by = "cmte_id")

# you can also write this as:
combined <- transactions %>% 
  left_join(committees, by = "cmte_id")
```

The left join is used here because our main table is `transactions`; we only care about information from `committees` here it has matches in `transactions` (i.e. there are a lot of committees that didn't report transactions in our subset, and we don't care about those.)

```{r}
# which pacs are giving the most to influence this race? 
combined %>% 
  group_by(cmte_nm) %>% 
  summarise(total = sum(transaction_amt)) %>% 
  arrange(desc(total))

# how much money in total? 
combined %>% 
  summarise(total = sum(transaction_amt))
```

This data is a small slice of the full data available from the FEC; I pulled only certain candidates related to our Missouri state elections in 2022. (I also added a few columns from the `candidates` table to avoid having to do multiple joins):

```{r}
# these are the columns that are originally from the `committees` table, which I joined to `transactions` using `cand_id`:
transactions %>% 
  count(cand_name, cand_pty_affil, cand_office)
```

Note the transaction types in this dataset: 
```{r}
transactions %>% 
  count(transaction_tp)

# 24A is an independent expenditure opposing a candidate
# 24E is an independent expenditure advocating for a candidate
# 24K is a committee to committee contribution
# 24Z is an in-kind contribution
```

Let's look at all the independent expenditures opposing candidates, and see which candidates received the most "negative" attention: 

```{r}
combined %>% 
  filter(transaction_tp == "24A") %>% 
  group_by(cand_name, cand_pty_affil, cand_office) %>% 
  summarise(total = sum(transaction_amt)) %>% 
  arrange(desc(total))

# Eric Greitens (our former governor who was running for Blunt's senate seat), by far.
```

It's worth noting that a join doesn't have to be the first thing you do. You can create a results table and then join that to something else: 

```{r}
# using transactions, find the committee that spent the most:
transactions %>% 
  group_by(cmte_id) %>%
  summarise(total = sum(transaction_amt)) %>% 
  arrange(desc(total))

# then join that results table to `committees`:
transactions %>% 
  group_by(cmte_id) %>%
  summarise(total = sum(transaction_amt)) %>% 
  arrange(desc(total)) %>% 
  left_join(committees, by = "cmte_id")

# this just demonstrates the flexibility of R / tidyverse.
```

Testing other types of joins to see what the results look like: 

Right join
```{r}
transactions %>% 
  right_join(committees, by = "cmte_id") %>% 
  filter(is.na(amndt_ind))
```

Inner join
```{r}
transactions %>% 
  inner_join(committees, by = "cmte_id")
```

Full join
```{r}
transactions %>% 
  full_join(committees, by = "cmte_id")
```

Semi join
```{r}
transactions %>% 
  semi_join(committees, by = "cmte_id")
```

Anti join
```{r}
transactions %>% 
  anti_join(committees, by = "cmte_id")
```




